---
title: "Reference-free Simulation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 4.8
)
library(STCOS)
```

This example builds a reference-free tissue with three cell types, 150 genes,
and 492 hexagonally arranged spots. It combines hotspots, streaks, layers,
cell-type-specific gene modules, and spatially varying cell-type composition.
Only ST-COS and base R are required.

<div class="note">
For image-derived layouts, open the
<a href="https://yiyanz.shinyapps.io/st-cos/">image-guided Shiny app</a>
directly or run <code>launch_stcos_app()</code> from R. The app accepts a
histology image, an externally segmented mask, or an uploaded coordinate file.
</div>

## 1. Define genes and spatial patterns

The first 50 genes carry designed spatial signal and the remaining 100 genes
have spatially constant means. The pattern assigned to each gene can differ by
cell type.

```{r parameters}
set.seed(1)

n_genes <- 150
genes <- paste0("gene", seq_len(n_genes))
cell_types <- c("ct1", "ct2", "ct3")

gene_pattern <- data.frame(
  ct1 = c(rep("hotspot", 10), rep("streak", 10),
          rep("no_pattern", 130)),
  ct2 = c(rep("no_pattern", 10), rep("streak", 10),
          rep("hotspot", 10), rep("no_pattern", 120)),
  ct3 = c(rep("no_pattern", 30), rep("hotspot", 10),
          rep("layer", 10), rep("no_pattern", 100)),
  row.names = genes
)

gene_relationship <- data.frame(
  ct1 = c(rep(NA, 10), rep(1, 3), rep(2, 3), rep(3, 4), rep(NA, 130)),
  ct2 = c(rep(NA, 10), rep(1, 3), rep(2, 3), rep(3, 4), rep(NA, 130)),
  ct3 = c(rep(NA, 40), rep(1, 10), rep(NA, 100)),
  row.names = genes
)

hotspot_centers <- list(
  ct1 = data.frame(
    x_coord = rep(25, 10),
    y_coord = c(rep(25, 3), rep(50, 4), rep(75, 3))
  ),
  ct2 = data.frame(
    x_coord = c(rep(25, 4), rep(75, 6)),
    y_coord = c(rep(75, 7), rep(25, 3))
  ),
  ct3 = data.frame(
    x_coord = c(rep(25, 3), rep(50, 4), rep(75, 3)),
    y_coord = rep(75, 10)
  )
)
```

`gene_relationship` defines three dependence blocks for the streak genes in
`ct1` and `ct2`, and one block for the layer genes in `ct3`. An `NA` leaves a
gene outside the explicitly correlated blocks.

## 2. Generate coordinates and layers

```{r coordinates}
coord <- generate_coordinates(
  x_len = 100,
  y_len = 100,
  pattern = "hex",
  spot_distance = 5
)
rownames(coord) <- paste0("spot_", seq_len(nrow(coord)))

layer_coord <- generate_layer(
  coord,
  xmin = c(0, 0, 50, 50),
  xmax = c(50, 50, 100, 100),
  ymin = c(0, 50, 0, 50),
  ymax = c(50, 100, 50, 100)
)

coord_key <- paste(coord$x_coord, coord$y_coord)
layer_key <- paste(layer_coord$x_coord, layer_coord$y_coord)
layer_coord <- layer_coord[match(coord_key, layer_key), ]
rownames(layer_coord) <- rownames(coord)

cat(sprintf(
  "%d spots, %d genes, %d cell types\n",
  nrow(coord), n_genes, length(cell_types)
))
```

## 3. Generate cell-type-specific mean fields

```{r mean-fields}
mean_param <- generate_gene_pattern(
  coord = coord,
  gene_pattern = gene_pattern,
  center_coord = hotspot_centers,
  layer_assignments = layer_coord$layer,
  layer_param = c(20, 55, 90, 130),
  seed = 123
)

# Preserve the separated streak positions used in the original example.
streak_centers <- list(
  ct1 = c(rep(10, 3), rep(20, 3), rep(30, 4)),
  ct2 = c(rep(60, 3), rep(70, 3), rep(80, 4))
)

for (cell_type in names(streak_centers)) {
  streak_genes <- which(gene_pattern[[cell_type]] == "streak")
  for (i in seq_along(streak_genes)) {
    mean_param[[cell_type]][, streak_genes[i]] <- generate_pattern(
      coord = coord,
      pattern = "streak",
      streak_x = streak_centers[[cell_type]][i],
      streak_x_param = 90,
      background_param = 20,
      streak_size = 18
    )
  }
}
```

```{r plot-means}
plot_field <- function(values, title) {
  palette <- hcl.colors(100, "YlOrRd", rev = TRUE)
  index <- cut(values, breaks = 100, include.lowest = TRUE, labels = FALSE)
  plot(
    coord$x_coord, coord$y_coord,
    pch = 21, bg = palette[index], col = NA,
    asp = 1, xlab = "x", ylab = "y", main = title
  )
}

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
plot_field(mean_param$ct1[, "gene1"], "ct1: hotspot")
plot_field(mean_param$ct2[, "gene11"], "ct2: streak")
plot_field(mean_param$ct3[, "gene45"], "ct3: layer")
```

## 4. Generate cell-type-specific expression

```{r expression}
set.seed(124)
cov_gene <- generate_covariance_group(gene_relationship)

cell_type_profiles <- generate_true_count(
  coord = coord,
  mean_param = mean_param,
  gene_relationship = gene_relationship,
  cov_gene = cov_gene,
  theta_g = setNames(
    rep(list(rep(10, n_genes)), length(cell_types)),
    cell_types
  ),
  eigen_floor = 1e-6
)

cat(
  sprintf(
    "%s: %d spots x %d genes",
    names(cell_type_profiles),
    vapply(cell_type_profiles, nrow, integer(1)),
    vapply(cell_type_profiles, ncol, integer(1))
  ),
  sep = "\n"
)
```

The covariance arrays define latent Gaussian dependence. Realized Pearson and
Spearman correlations in the generated counts can be lower because the
negative-binomial inverse CDF is discrete and gene marginals differ.

## 5. Define spatial cell-type composition

Each quadrant has a different expected composition. Small random variation is
added within each layer, and every row is normalized to sum to one.

```{r composition}
composition_by_layer <- data.frame(
  ct1 = c(1, 1 / 3, 0, 0),
  ct2 = c(0, 1 / 3, 1 / 2, 4 / 5),
  ct3 = c(0, 1 / 3, 1 / 2, 1 / 5)
)

celltype_proportion <- generate_cell_prop(
  layer_coord = layer_coord,
  layer_param = composition_by_layer
)
rownames(celltype_proportion) <- rownames(coord)

round(head(celltype_proportion), 3)
range(rowSums(celltype_proportion))
```

```{r plot-composition}
par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
for (cell_type in cell_types) {
  plot_field(
    celltype_proportion[, cell_type],
    paste(cell_type, "proportion")
  )
}
```

## 6. Construct the spot-level expression matrix

```{r spot-expression}
spot_expression <- generate_spot_count(
  true_count = cell_type_profiles,
  nGenes = n_genes,
  cell_number = 10,
  celltype_proportion = celltype_proportion
)

dim(spot_expression)
spot_expression[1:4, 1:6]
```

Rows are spots and columns are genes. `cell_number` is the total
cell-equivalent abundance used by the abundance-weighted spot mixture, so the
resulting values need not be integers.

## 7. Inspect the dependence structure

```{r correlations}
plot_correlation <- function(x, title) {
  correlation <- cor(x[, 1:50], method = "spearman")
  image(
    seq_len(ncol(correlation)), seq_len(nrow(correlation)),
    t(correlation[nrow(correlation):1, ]),
    col = hcl.colors(101, "Blue-Red 3"), zlim = c(-1, 1),
    axes = FALSE, xlab = "Genes 1-50", ylab = "Genes 1-50",
    main = title
  )
  box()
}

par(mfrow = c(2, 2), mar = c(3, 3, 3, 1))
plot_correlation(cell_type_profiles$ct1, "ct1")
plot_correlation(cell_type_profiles$ct2, "ct2")
plot_correlation(cell_type_profiles$ct3, "ct3")
plot_correlation(spot_expression, "Spot mixture")
```

## Adapt the design

- Edit `gene_pattern` to move a hotspot, streak, or layer signal to another
  cell type.
- Edit one block of `gene_relationship` or `cov_gene` to create a controlled
  co-expression intervention.
- Replace `composition_by_layer` with another designed or externally supplied
  abundance field.
- Pass a vector to `cell_number` to vary total abundance among spots.
- Use `generate_batch_count()` to add dropout and sequencing-depth effects.

For fitted simulations, continue to the
[reference-based tutorial](Data_code.html).
