---
title: "Reference-free Simulation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 4.5
)
library(STCOS)
```

This tutorial builds a small simulation with two cell types, four genes,
cell-type-specific spatial patterns, and one dependence block per cell type.
Only ST-COS and base R are required.

## 1. Generate coordinates

```{r coordinates}
set.seed(1)

coord <- generate_coordinates(
  x_len = 10,
  y_len = 10,
  pattern = "grid",
  spot_distance = 1
)
rownames(coord) <- paste0("spot_", seq_len(nrow(coord)))

dim(coord)
head(coord)
```

`pattern` can be `"grid"`, `"hex"`, or `"random"`. For random coordinates,
provide `n_spot` instead of `spot_distance`.

## 2. Specify spatial mean patterns

Rows of `gene_pattern` are genes and columns are cell types. Available labels
are `"hotspot"`, `"streak"`, `"layer"`, and `"no_pattern"`.

```{r mean-patterns}
genes <- paste0("gene_", 1:4)

gene_pattern <- data.frame(
  Excitatory = c("hotspot", "hotspot", "no_pattern", "no_pattern"),
  Inhibitory = c("no_pattern", "no_pattern", "streak", "streak"),
  row.names = genes
)

mean_param <- generate_gene_pattern(
  coord = coord,
  gene_pattern = gene_pattern,
  seed = 1
)

lapply(mean_param, dim)
```

Each element of `mean_param` is a spot-by-gene mean matrix. The optional
pattern arguments let users set hotspot centers and radii, streak positions
and widths, or per-layer means.

```{r plot-means}
plot_field <- function(values, title) {
  palette <- hcl.colors(100, "YlOrRd", rev = TRUE)
  index <- cut(values, breaks = 100, include.lowest = TRUE, labels = FALSE)
  plot(
    coord$x_coord, coord$y_coord,
    pch = 21, bg = palette[index], col = NA,
    asp = 1, xlab = "x", ylab = "y", main = title
  )
}

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot_field(mean_param$Excitatory[, "gene_1"], "Excitatory: gene_1")
plot_field(mean_param$Inhibitory[, "gene_3"], "Inhibitory: gene_3")
```

## 3. Specify latent gene dependence

`gene_relationship` identifies the genes included in each dependence block.
`NA` marks genes simulated independently in that cell type.

```{r dependence}
gene_relationship <- data.frame(
  Excitatory = c(1, 1, NA, NA),
  Inhibitory = c(NA, NA, 1, 1),
  row.names = genes
)

rho <- array(
  0,
  dim = c(4, 4, 2),
  dimnames = list(genes, genes, c("Excitatory", "Inhibitory"))
)
diag(rho[, , "Excitatory"]) <- 1
diag(rho[, , "Inhibitory"]) <- 1

rho[1, 2, "Excitatory"] <- rho[2, 1, "Excitatory"] <- 0.6
rho[3, 4, "Inhibitory"] <- rho[4, 3, "Inhibitory"] <- 0.6
```

The value `0.6` is a **latent Gaussian correlation**, not a guarantee that the
final count-scale Pearson correlation will equal 0.6.

## 4. Generate cell-type-specific profiles

```{r cell-type-profiles}
cell_type_profiles <- generate_true_count(
  coord = coord,
  mean_param = mean_param,
  gene_relationship = gene_relationship,
  cov_gene = rho,
  theta_g = list(
    Excitatory = rep(10, length(genes)),
    Inhibitory = rep(10, length(genes))
  ),
  eigen_floor = 1e-6
)

lapply(cell_type_profiles, dim)
cell_type_profiles$Excitatory[1:4, ]
```

Larger negative-binomial size values imply lower overdispersion at a fixed
mean. A separate value can be supplied for every gene and cell type.

## 5. Define cell-type composition

Here, excitatory abundance increases smoothly from left to right.

```{r composition}
p_excitatory <- plogis((coord$x_coord - 5) / 1.5)
celltype_proportion <- cbind(
  Excitatory = p_excitatory,
  Inhibitory = 1 - p_excitatory
)
rownames(celltype_proportion) <- rownames(coord)

range(rowSums(celltype_proportion))
head(celltype_proportion)
```

```{r plot-composition}
par(mfrow = c(1, 1), mar = c(4, 4, 3, 1))
plot_field(celltype_proportion[, "Excitatory"], "Excitatory proportion")
```

## 6. Construct spot-level expression

```{r spot-expression}
spot_expression <- generate_spot_count(
  true_count = cell_type_profiles,
  nGenes = length(genes),
  cell_number = 8,
  celltype_proportion = celltype_proportion
)

dim(spot_expression)
spot_expression[1:4, ]
```

The rows are spots and the columns are genes. `cell_number` is interpreted as
the total cell-equivalent abundance. Because the construction is a continuous
abundance-weighted mixture, the resulting spot values need not be integers.

## 7. Inspect the requested and realized dependence

```{r inspect-correlation}
requested <- rho[, , "Excitatory"]
realized_pearson <- cor(cell_type_profiles$Excitatory, method = "pearson")
realized_spearman <- cor(cell_type_profiles$Excitatory, method = "spearman")

round(c(
  latent = requested[1, 2],
  count_pearson = realized_pearson[1, 2],
  count_spearman = realized_spearman[1, 2]
), 3)
```

Differences between the requested latent value and count-scale estimates are
expected. They depend on the means, dispersions, zero frequencies, spatial
mean patterns, and aggregation.

## Customize the scenario

- Change `gene_pattern` to move spatial signal among genes or cell types.
- Edit one block of `rho` to create a controlled dependence intervention.
- Replace `celltype_proportion` with an externally generated abundance field.
- Pass a vector to `cell_number` to vary total abundance by spot.
- Use `generate_batch_count()` to add dropout and sequencing-depth effects.

For fitted simulations, continue to the
[reference-based tutorial](Data_code.html).
