---
title: "Installation and Inputs"
output: html_document
---

## Install ST-COS

Install the development version directly from GitHub:

```{r installation, eval=FALSE}
install.packages("devtools") # Skip if devtools is already installed
devtools::install_github("Yiyannnnn/ST-COS")
```

Load the package and confirm the installed version:

```{r load-package, eval=FALSE}
library(STCOS)
packageVersion("STCOS")
```

The package imports `data.table`, `expm`, `mgcv`, and `spdep`. Installing with
`devtools::install_github()` installs missing CRAN dependencies automatically.

## Launch the image-guided application

The hosted Shiny application accepts histology images or coordinate files and
provides interactive controls for spatial layout, regions, gene patterns,
cell-type composition, and gene-dependence groups.

```{r launch-app, eval=FALSE}
launch_stcos_app()
```

This opens [the ST-COS image-guided app](https://yiyanz.shinyapps.io/st-cos/)
in the default browser. From a remote or headless R session, return the URL
without opening a browser:

```{r app-url, eval=FALSE}
launch_stcos_app(browser = FALSE)
```

## Input conventions

| Object | Orientation and requirements |
| --- | --- |
| Reference count matrix | Genes in rows and spots in columns; non-negative counts |
| Reference coordinates | One row per spot with numeric columns named `row` and `col` |
| Reference-free coordinates | One row per spot with columns `x_coord` and `y_coord` |
| `mean_param` | Named list of spot-by-gene matrices, one per cell type |
| `theta_g` | Named list containing one NB size per gene, or one shared value |
| `celltype_proportion` | Spot-by-cell-type matrix; every row sums to one |
| `cov_gene` | Gene-by-gene matrix or array with one matrix per cell type |

Cell-type names must agree across `mean_param`, `theta_g`,
`gene_relationship`, `cov_gene`, and the columns of `celltype_proportion`.

## Main functions

| Function | Purpose |
| --- | --- |
| `generate_coordinates()` | Create grid, hexagonal, or random locations |
| `generate_gene_pattern()` | Build cell-type-specific spatial mean matrices |
| `generate_covariance_group()` | Generate grouped dependence matrices |
| `generate_true_count()` | Simulate cell-type-specific spot profiles |
| `generate_spot_count()` | Construct abundance-weighted spot profiles |
| `generate_batch_count()` | Apply dropout and sequencing-depth effects |
| `get_real_param()` | Estimate reference-based parameters |
| `launch_stcos_app()` | Open the hosted image-guided interface |

Use `?function_name` in R for argument-level documentation, for example:

```{r help, eval=FALSE}
?generate_true_count
?get_real_param
```

## Reviewed defaults

The reference-based function exposes the sensitivity parameters used in the
manuscript:

```{r defaults, eval=FALSE}
fitted <- get_real_param(
  expr,
  spatial_coord,
  rownames(expr),
  moran_thr = 0.15,
  cor_thr = 0.05,
  eigen_floor = 1e-6,
  neighbor_k = 6,
  spline_k = 50,
  dispersion_cap = 1e6,
  n_cores = 4
)
```

- `eigen_floor` is a numerical device; routine searching is not required.
- `cor_thr` is substantive because it controls sparsification of the fitted
  dependence matrix.
- `neighbor_k = 6` is the default used for the Visium analyses in the study,
  but it remains user-configurable for other spatial resolutions.
- `dispersion_cap = Inf` requests the explicit Poisson limit when the estimated
  variance does not exceed the mean.

## Quick installation check

This small check should return a coordinate table with 25 rows:

```{r install-check, eval=FALSE}
coord <- generate_coordinates(
  x_len = 4,
  y_len = 4,
  pattern = "grid",
  spot_distance = 1
)
dim(coord)
# [1] 25  2
```

Next: [run a reference-free simulation](Example.html).
