---
title: "Interpretation and Limitations"
output: html_document
---

## Latent and count-scale correlation are different

ST-COS uses a Gaussian copula. The supplied matrix controls Pearson
correlation among latent Gaussian variables. After transformation through
gene-specific negative-binomial inverse CDFs,

$$
W_g \longrightarrow \Phi(W_g) \longrightarrow
F^{-1}_{\mathrm{NB},g}\{\Phi(W_g)\},
$$

the realized count-scale Pearson correlation generally differs from the
latent target. Spearman correlation can also change because count data contain
many ties.

Attenuation is usually stronger with sparse or highly dispersed marginals.
Extra zeros can alter Pearson and Spearman correlation differently. For this
reason, report the requested latent value and the realized count-scale value
separately.

```{r correlation-check, eval=FALSE}
latent_target <- fitted$cov_gene[, , "CT"]
count_pearson <- cor(simulated_counts, method = "pearson")
count_spearman <- cor(simulated_counts, method = "spearman")
```

ST-COS does not currently invert the copula parameter to hit an exact
count-scale Pearson target. Such calibration would depend on each gene's mean,
dispersion, and zero frequency.

## What the dependence model represents

The Gaussian copula provides editable **pairwise** dependence. It does not
claim to reproduce arbitrary higher-order or tail dependence. Use the terms
below consistently:

| Term | Meaning in ST-COS |
| --- | --- |
| Latent correlation | Pearson correlation in the Gaussian copula |
| Count Pearson correlation | Pearson correlation measured from simulated counts |
| Count Spearman correlation | Rank correlation measured from simulated counts |
| Co-expression edge | A thresholded or ranked relationship defined by an evaluation rule |

## Reference-based scope

Reference-based mode estimates one global spot-level dependence matrix. In a
heterogeneous tissue, this matrix can reflect both within-cell-type expression
and changing cell-type composition. The current release does not deconvolve
these sources when estimating dependence.

Use one of the following when cell-type-specific dependence is required:

1. Supply externally estimated cell-type-specific matrices.
2. Specify the matrices directly in reference-free mode.
3. Use a cell-resolved reference and construct each cell-type matrix outside
   the current reference-based estimator.

## Spot-level abundance weighting

`generate_spot_count()` computes

$$
X_{jg}=\sum_c N_jp_{jc}Y_{jg}^{(c)}.
$$

Here, $Y_{jg}^{(c)}$ is a latent average expression level for cell type $c$ at
spot $j$. The model does not draw and sum $N_jp_{jc}$ independent individual
cells. Consequently, `cell_number` represents cell-equivalent abundance, and
the resulting weighted spot profile can contain non-integer values.

This formulation is useful for controlled spot-level mixtures. It should not
be described as an explicit model of within-spot cell-to-cell variability.

## Positive-definite repair

Thresholding a correlation matrix can make it non-positive-definite. ST-COS
repairs the matrix before Cholesky factorization using `eigen_floor`, whose
default is `1e-6`.

Reference fitting returns these diagnostics:

- minimum eigenvalue before and after repair;
- minimum eigenvalue lift;
- number of modified eigenvalues;
- relative Frobenius perturbation; and
- maximum entry-wise change.

```{r matrix-diagnostics, eval=FALSE}
fitted$matrix_diagnostics
```

The floor is a numerical safeguard. By contrast, `cor_thr` controls
sparsification and can materially change the requested matrix.

## Reporting checklist

For a reproducible ST-COS experiment, report:

- package version and random seed;
- number of spots, genes, and cell types;
- coordinate and composition construction;
- negative-binomial means and sizes;
- requested latent dependence structure;
- realized Pearson and Spearman correlations;
- `cor_thr`, `eigen_floor`, `neighbor_k`, `spline_k`, and `dispersion_cap` when
  reference-based fitting is used; and
- matrix-repair diagnostics and runtime environment.

For benchmark claims, use independent replicates and uncertainty intervals,
and avoid ranking language when methods are separated only by near-ties.

Return to the [tutorial overview](index.html).
